Towards Detecting Stalkerware Through Network Traces

### 2019-03-23 paper/main.tex
Minor edits

\begin{abstract}
Intimate partner violence (IPV) is a pervasive problem that has been changed by 
the advent of the digital age. The ubiquity of smartphones has made it easy 
even for someone with no technical skills to perform surveillance of a device 
that they have access to. In the context of IPV, abusers often have access to 
the devices of their partners, and either know or can coerce the passwords to 
those devices. With such unrestricted access, an abundance of surveillance 
options exist, 
whether the goal is to track location, remotely block incoming calls or 
messages, record all communications, monitor browsing history, or lock a 
survivor out of their accounts. For some types of information, no extra 
software is required to gather it. For example, Facebook can reveal a user's 
location to their friends. Another example is Apple's backup service, which can 
make copies of text messages and call logs accessible to anyone with the iCloud 
credentials of the phone. Other types of control that might be attractive to an 
abuser, such as the ability to block incoming communication from certain 
people, require software to be installed on the target device. We refer to such 
software as ``stalkerware.'' Stalkerware is dissimilar to other forms of 
malware because of how easily the adversary can access the target device. 
Consequently, traditional assumptions about security do not apply to it, and 
traditional methods of studying it are ineffective. Antivirus, which can 
normally provide information about malware, is poor at detecting 
stalkerware.\cite{chatterjee_spyware_2018} Network Intrusion Detection Systems 
(IDSs) also have low detection rates. This lack of information results in many 
unanswered questions about the use of stalkerware in the context of IPV.

This work takes a step towards answering one of those questions: namely, 
\textit{how prevalent is stalkerware today?} We identify the traffic signatures 
of 24 stalkerware apps, with the goal of being able to identify it from a 
network operator's vantage point. We show that our signatures identify 
stalkerware with few false positives. Additionally, we present a simple method 
of generalizing the signatures we identify, and show that our generalized 
signatures also give few false positives. We also show that peculiarities of 
the stalkerware ecosystem may aid researchers in detecting it, and in 
generalizing specific network signatures to apps that have not been manually 
identified. The ethical and bureaucratic issues inherent in doing a large-scale 
measurement study on real traffic prevented us from using our signatures to 
estimate how much stalkerware exists in the wild. However, we present a plan 
for how we could move forward. 
\end{abstract}

@@

\section{Introduction}
Intimate partner violence, or IPV, is a pervasive problem in the United States 
and elsewhere. One in four women and one in six men will be subject to some 
sort of intimate 
partner abuse during their lifetimes, where ``intimate partner abuse'' can 
include physical, emotional, or sexual abuse, as well as stalking or rape. In a 
world where technology now pervades every aspect of our lives, some 
aspects of IPV have changed too. Technology can now be abused in a variety of 
ways that harm survivors of IPV. Previous work \cite{ristenpart_ucsd_talk} has 
classified these into four categories of attacks: ownership-based, account or 
device compromise, harmful messages or posts, and exposure of private 
information. Ownership-based attacks are possible when the abuser owns the 
device used by the victim or the victim's children. They are then capable 
of installing software on it that can remotely destroy the device or prevent or 
monitor access and usage. Attacks based on account or device compromise are 
possible when the abuser does not own the device but nevertheless can guess 
passwords or compel the victim to reveal them, or has access to the unlocked 
device, or can remotely ``hack'' it using knowledge of security questions or 
passwords. Some examples of a ``harmful message'' attack include messaging the 
victim from a spoofed account, messaging the victim's family or friends, or 
impersonating the victim using their own accounts. Finally, attacks that expose 
private information can involve ``revenge pornography,'' revealing an HIV 
status, creating fake profiles offering sexual services, and blackmail. While 
some of these options (such as tracking a victim's location) are available to 
an abuser without installing any extra software on the phone, others are not. 
We refer to monitoring software that can record online activity, 
communications, and location, or remotely tamper with the device, as 
``stalkerware.''

Stalkerware is not as widely studied as other types of malware, and much less 
information about it is available. There are a few key differences between it 
and other forms of malware. First, stalkerware is meant to be installed 
by someone with physical access to the target device, not remotely by 
exploiting vulnerabilities. In the 
context of intimate partner abuse, this presents an enormous 
threat, given how frequently the abuser has unfettered access to the target 
phone. Second, the 
threat model used by malware developers includes sophisticated antivirus and 
intrusion detection systems (IDSs). Both malware and stalkerware need to hide 
their presence on the target device, but malware is much more likely to 
attempt to obfuscate its network traffic to thwart these adversaries. 
Stalkerware does not: it does not 
need to. Most antivirus does not detect apps capable of monitoring a target as 
potentially dangerous \cite{chatterjee_spyware_2018}. This presents an 
opportunity to glean information about stalkerware that is not available in the 
study of other forms of malware. At the same time, it limits a researcher's 
ability to use information collected by antivirus as a source of information 
about stalkerware. Third, stalkerware falls 
for the most part into two categories: apps that are explicitly advertised for 
surveilling spouses and partners, and apps that have a different stated use but 
can be repurposed to stalk a target. These are usually marketed for theft 
prevention, finding lost devices, parental control of children, and employer 
control of employees. Previous literature refers to these as "dual-use" 
apps\cite{chatterjee_spyware_2018}. Some dual-use apps are ones that the 
victim wants to keep installed, such as 
``find my phone'' apps or social media apps that leak the user's location. The 
most troubling example of this is Facebook, which most victims want to keep 
so that they can contact their support network for help.  Determining if a 
particular dual-use app is being used for intimate partner surveillance (IPS) 
presents an enormous challenge. Given the wide spectrum of apps that can be 
used to facilitate IPS and the limited scope of this project, we chose to focus 
on more overt spyware.

Largely due to the differences between stalkerware and other malware, most 
available information on the \textit{pervasiveness} of stalkerware is anecdotal 
and piecemeal. Small-scale studies on how widespread spyware is do exist, but 
they have limitations. For example, in a survey of 72 
domestic violence shelters performed by NPR in 2014, 85\% said they worked with 
survivors whose abusers had tracked their location using their smartphone's 
GPS. 75\% said they have worked with clients who have had their conversations 
monitored by hidden stalkerware apps.\cite{shahani_smartphones_nodate} However, 
this study does not report how many individual clients experienced being 
digitally stalked. Additionally, it and similar work relies on 
reports from domestic violence shelters and health workers, which do not give 
a complete picture of the situation. Little to no data exist regarding spyware 
use on IPV victims who never visit a shelter. Traditionally, antivirus vendors 
collect metrics about 
malicious software, but antivirus often does not detect spyware 
\cite{chatterjee_spyware_2018}. Alarmingly, the stalkerware 
websites themselves claim to have hundreds of thousands of downloads per app, 
but these numbers are not likely to be trustworthy. The goal of this work is 
therefore to identify a means of identifying stalkerware in the wild so that it 
can be enumerated. We identified unique signatures in the traffic of 24 
stalkerware apps, by examining a traffic features such as domain names, HTTP 
header fields, protocol, and server name indicators (SNIs). For a complete 
list, see section \ref{network_signatures}. These signatures can be used by 
antivirus software and IDSs to determine how widespread stalkerware is. We show 
that these signatures are distinguishable from ordinary smartphone 
network traffic. We also propose a simple method of generalizing our signatures 
to spyware that we did not manually examine, and evaluate its robustness.

We began by identifying a set of apps that appear to be 
marketed primarily for IPS. We focused exclusively on Android apps, since our 
two test phones were both Androids. Apps that appear in Google searches for 
terms like 
"catch a cheating spouse," use vague language that appear to be marketed 
towards IPS, and label themselves as "spyware" became candidates that we 
considered. Overt spyware is not generally available from the app store - it 
must be downloaded from the manufacturer sites. It also costs much more than a 
typical app, with monthly 
subscriptions ranging from \$12 to \$70 per month. The exception is an app that 
we call SpyToMobile, which survives on the app store by frequently 
``reskinning'' itself (see section \ref{reskinning}).  
We selected several of overt apps to test. All of these apps brand themselves 
as spyware and advertise that their products are undetectable on the target 
phone, despite also having disclaimers in small text that their software should 
not be used for any illegal activity. Some use language that appears designed 
to appeal to abusers. For example, underneath one promotional 
video, the app TheOneSpy has the disclaimer: ``Note for this video: Don't 
confuse with the world `Loved Ones' as spouse or Girl friends! Word Loved ones 
has used for kids and teens.''\cite{noauthor_theonespy_2013} All of these apps 
advertise that once they are installed, they are in ``stealth mode'' and are 
completely undetectable. This is not entirely the case - they appear in the 
list of all applications, but take care to name themselves something innocuous 
such as ``Data Backup'' or ``System Service.'' None of the overt apps that we 
examined required rooting the phone. When installed, they ask for all the 
permissions they 
require, including device administrator privileges. They are then capable of 
blocking incoming calls, recording messages that have been deleted, recording 
the screen of the phone using accessibility features, monitoring apps such as 
Facebook, WhatsApp, Kik, and Viber, and more.

Other apps were explicitly intended as spyware, but nevertheless available from 
the Play Store. They fell into two categories: spyware that reskinned itself to 
appear to be many different apps, and apps designed for parents to keep track 
of children (or in one case, for adult children to track elderly parents.) For 
the most part, this latter category was offered by antivirus companies such as 
Avast and Norton. These apps offered 
multiple features, such as call screening, blocking Google search results, 
recording text messages and call logs, and ``panic buttons'' that would allow 
the surveilled phone to contact the surveillant and raise an alarm. They do not 
hide their app icons and generally make an effort to notify the user that the 
monitoring app is installed on their phone. The spyware that we found on the 
Google Play store that did \textit{not} notify users of its presence consisted 
of one app, which we call SpyToMobile after the title on its loading screen, 
that was provided on the Play Store under many different names (see section 
\ref{reskinning}).

We also identified many apps that were designed to track location or text 
messages, but were less comprehensive than the overt spyware. They 
track one feature each instead of tracking location, communication, and 
browsing history all at once. For the most part, these apps were free and 
available on the Google Play Store. Most of them were designed to keep track 
of the locations of the users' friends. They tended to be ineffective - nearly 
all of the apps we found that did not work fell into this category. Most of 
these apps were unable or unwilling to hide their icons, and some provide an 
explicit notification that the app is in use, although this notification can be 
easily disabled in the Android notification settings.

\section{Related Work}
Previous studies \cite{stories_from_survivors, burke_using_2011}, the media 
\cite{shahani_smartphones_nodate, chatterjee_spyware_2018}, and intimate 
partner violence (IPV) survivors \cite{ristenpart_ucsd_talk} have all reported 
that spyware is becoming an increasingly utilized means of performing
intimate partner surveillance (IPS). Tracking functions supported by spyware 
allow stalkers to monitor communications, track locations, and remotely 
activate cameras and microphones \cite{shahani_smartphones_nodate, 
chatterjee_spyware_2018}. Chatterjee et al. propose methods for searching for 
IPS-capable apps. By Google search query expansion and unsupervised 
classification, they collected 9,224 apps possibly related to IPS on Google 
play and the open Internet. They manually selected 70 apps for investigation. 
They observed the most fundamental functions of these apps are 
monitoring locations, communication logs and data, media contents, and phone 
usages. They also found that existing anti-spyware tools cannot effectively 
detect IPS-related apps \cite{chatterjee_spyware_2018}.
Mobile apps are often identified via the User-Agent string of the HTTP request, 
when one is available \cite{xu_identifying_2011}. With the increased use of 
HTTPS this may not always 
be the case, but we expect to find (based on conversations with authors of 
previous work) that security and best coding practices are not prioritized by 
stalkerware developers. Malware identification may be of use as well, 
since techniques are available to cluster malware that comes from the same 
source code into "families." \cite{hutchison_firma_2013}

>>>>

\begin{abstract}
Intimate partner violence (IPV) is a pervasive problem that has been changed by 
the advent of the digital age. The ubiquity of smartphones has made it easy 
even for someone with no technical skills to perform surveillance of a device 
that they have access to. In the context of IPV, abusers often have access to 
the devices of their partners, and either know or can coerce the passwords to 
those devices. With such unrestricted access, an abundance of surveillance 
options exist, 
whether the goal is to track location, remotely block incoming calls or 
messages, record all communications, monitor browsing history, or lock a 
survivor out of their accounts. No extra software is required to gather some 
types of information. For example, Facebook can reveal a user's 
location to their friends. Another example is Apple's backup service, which can 
make copies of text messages and call logs accessible to anyone with the iCloud 
credentials of the phone. Other types of control that might be attractive to an 
abuser, such as the ability to block incoming communication from certain 
people, require software to be installed on the target device. We refer to such 
software as ``stalkerware.'' Stalkerware is dissimilar to other forms of 
malware because of how easily the adversary can access the target device. 
Consequently, traditional assumptions about security do not apply to it, and 
traditional methods of studying it are ineffective. Antivirus, which can 
normally provide information about malware, is poor at detecting 
stalkerware.\cite{chatterjee_spyware_2018} Network Intrusion Detection Systems 
(IDSs) also have low detection rates. This lack of information results in many 
unanswered questions about the use of stalkerware in the context of IPV.

This work takes a step towards answering one of those questions: namely, 
\textit{how prevalent is stalkerware today?} We identify the traffic signatures 
of 24 stalkerware apps, with the goal of being able to identify them from a 
network operator's vantage point. We show that our signatures identify 
stalkerware with few false positives. Additionally, we present a simple method 
of generalizing the signatures we identify, and show that our generalized 
signatures also give few false positives. We also show that peculiarities of 
the stalkerware ecosystem may aid researchers in detecting stalkerware, and in 
generalizing specific network signatures to apps that have not been manually 
identified. The ethical issues inherent in doing a large-scale 
measurement study on real traffic prevented us from using our signatures to 
estimate how much stalkerware exists in the wild. However, we present a plan 
for how we could move forward. 
\end{abstract}

@@

\section{Introduction}
Intimate partner violence, or IPV, is a pervasive problem in the United States 
and elsewhere. One in four women and one in six men will be subject to some 
sort of intimate 
partner abuse during their lifetimes, where ``intimate partner abuse'' can 
include physical, emotional, or sexual abuse, as well as stalking or rape. In a 
world where technology now pervades every aspect of our lives, some 
aspects of IPV have changed too. Technology can now be abused in a variety of 
ways that harm victims of IPV. Previous work \cite{ristenpart_ucsd_talk} has 
classified these into four categories of attacks: ownership-based attacks, 
account or device compromise, harmful messages or posts, and exposure of 
private 
information. Ownership-based attacks are possible when the abuser owns the 
device used by the victim or the victim's children. They are then capable 
of installing software on it that can remotely destroy the device or prevent or 
monitor access and usage. Attacks based on account or device compromise are 
possible when the abuser does not own the device but nevertheless can guess 
passwords or compel the victim to reveal them, or has access to the unlocked 
device, or can remotely ``hack'' it using knowledge of security questions or 
passwords. Some examples of a ``harmful message'' attack include messaging the 
victim from a spoofed account, messaging the victim's family or friends, or 
impersonating the victim using their own accounts. Finally, attacks that expose 
private information can involve ``revenge pornography,'' revealing an HIV 
status, creating fake profiles offering sexual services, and blackmail. While 
some of these options (such as tracking a victim's location) are available to 
an abuser without installing any extra software on the phone, others are not. 
We refer to monitoring software that can record online activity, 
communications, and location, or remotely tamper with the device, as 
``stalkerware.''

Stalkerware is not as widely studied as other types of malware, and much less 
information about it is available. There are a few key differences between it 
and other forms of malware. First, stalkerware is meant to be installed 
by someone with physical access to the target device, not remotely by 
exploiting vulnerabilities. In the 
context of intimate partner abuse, this presents an enormous 
threat, given how frequently the abuser has unfettered access to the target 
phone. Second, the 
threat model used by malware developers includes sophisticated antivirus and 
intrusion detection systems (IDSs). Both malware and stalkerware need to hide 
their presence on the target device, but malware is much more likely to 
attempt to obfuscate its network traffic to thwart these adversaries. 
Stalkerware does not: it does not 
need to. Most antivirus does not detect apps capable of monitoring a target as 
potentially dangerous \cite{chatterjee_spyware_2018}. This presents an 
opportunity to glean information about stalkerware that is not available in the 
study of other forms of malware. At the same time, it limits a researcher's 
ability to use information collected by antivirus as a source of information 
about stalkerware. Third, stalkerware falls 
for the most part into two categories: apps that are explicitly advertised for 
surveilling spouses and partners, and apps that have a different stated use but 
can be repurposed to stalk a target. These are usually marketed for theft 
prevention, finding lost devices, parental control of children, and employer 
control of employees. Previous literature refers to these as ``dual-use'' 
apps \cite{chatterjee_spyware_2018}. Some dual-use apps are ones that the 
victim wants to keep installed, such as 
``find my phone'' apps or social media apps that leak the user's location. The 
most troubling example of this is Facebook, which most victims want to keep 
so that they can contact their support network for help.  Determining if a 
particular dual-use app is being used for intimate partner surveillance (IPS) 
presents an enormous challenge. Given the wide spectrum of apps that can be 
used to facilitate IPS and the limited scope of this project, we chose to focus 
on more overt spyware.

Largely due to the differences between stalkerware and other malware, most 
available information on the \textit{pervasiveness} of stalkerware is anecdotal 
and piecemeal. Small-scale studies on how widespread spyware is do exist, but 
they have limitations. For example, in a survey of 72 
domestic violence shelters performed by NPR in 2014, 85\% said they worked with 
survivors whose abusers had tracked their location using their smartphone's 
GPS. 75\% said they have worked with clients who have had their conversations 
monitored by hidden stalkerware apps \cite{shahani_smartphones_nodate}. 
However, 
this study does not report how many individual clients experienced being 
digitally stalked. Additionally, it and similar work relies on 
reports from domestic violence shelters and health workers, which do not give 
a complete picture of the situation. Little to no data exist regarding spyware 
use on IPV victims who never visit a shelter. Traditionally, antivirus vendors 
collect metrics about 
malicious software, but antivirus often does not detect spyware 
\cite{chatterjee_spyware_2018}. Alarmingly, the stalkerware 
websites themselves claim to have hundreds of thousands of downloads per app, 
but these numbers are not likely to be trustworthy. The goal of this work is 
therefore to identify a means of identifying stalkerware in the wild so that it 
can be enumerated. We identified unique signatures in the traffic of 24 
stalkerware apps, by examining traffic features such as domain names, HTTP 
header fields, protocol, and server name indicators (SNIs). For a complete 
list, see section \ref{network_signatures}. These signatures can be used by 
antivirus software and IDSs to determine how widespread stalkerware is. We show 
that these signatures are distinguishable from ordinary smartphone 
network traffic. We also propose a simple method of generalizing our signatures 
to spyware that we did not manually examine, and evaluate its robustness.

We began by identifying a set of apps that appear to be 
marketed primarily for IPS. We focused exclusively on Android apps, since our 
two test phones were both Androids. Apps that appear in Google searches for 
terms like ``catch a cheating spouse,'' use vague language that appear to be 
marketed towards IPS, and label themselves as ``spyware'' became candidates 
that we considered. Stalkerware marketed for tracking a romantic partner is not 
generally available from the app store - it must be downloaded from the 
manufacturer sites. It also costs much more than a typical app, with monthly 
subscriptions ranging from \$12 to \$70 per month. The exception is an app that 
we call SpyToMobile, which survives on the app store by frequently 
``reskinning'' itself (see section \ref{reskinning}).  
We selected several of the most overt apps to test. All of these apps brand 
themselves 
as spyware and advertise that their products are undetectable on the target 
phone, despite also having disclaimers in small text that their software should 
not be used for any illegal activity. Some use language that appears designed 
to appeal to abusers. For example, underneath one promotional 
video, the app TheOneSpy has the disclaimer: ``Note for this video: Don't 
confuse with the world `Loved Ones' as spouse or Girl friends! Word Loved ones 
has used for kids and teens.''\cite{noauthor_theonespy_2013} All of these apps 
advertise that once they are installed, they are in ``stealth mode'' and are 
completely undetectable. This is not entirely the case - they appear in the 
list of all applications, but take care to name themselves something innocuous 
such as ``Data Backup'' or ``System Service.'' None of the overt apps that we 
examined required rooting the phone. When installed, they ask for all the 
permissions they 
require, including device administrator privileges. They are then capable of 
blocking incoming calls, recording messages that have been deleted, recording 
the screen of the phone using accessibility features, monitoring apps such as 
Facebook, WhatsApp, Kik, and Viber, and more.

Other apps were explicitly intended as spyware, but nevertheless available from 
the Play Store. They fell into two categories: spyware that reskinned itself to 
appear to be many different apps, and apps designed for parents to keep track 
of children (or in one case, for adult children to track elderly parents.) For 
the most part, this latter category was offered by antivirus companies such as 
Avast and Norton. These apps offered 
multiple features, such as call screening, blocking Google search results, 
recording text messages and call logs, and ``panic buttons'' that would allow 
the surveilled phone to contact the surveillant and raise an alarm. They do not 
hide their app icons and generally make an effort to notify the user that the 
monitoring app is installed on their phone. The spyware that we found on the 
Google Play store that did \textit{not} notify users of its presence mostly 
consisted of copies of one app, which we call SpyToMobile after the title on 
its loading screen, 
that was provided on the Play Store under many different names (see section 
\ref{reskinning}).

We also identified many apps that were designed to track location or text 
messages, but were less comprehensive than the overt spyware. They 
track one feature each instead of tracking location, communication, and 
browsing history all at once. For the most part, these apps were free and 
available on the Google Play Store. Most of them were designed to keep track 
of the locations of the users' friends. They tended to be ineffective - nearly 
all of the apps we found that did not work fell into this category. Most of 
these apps were unable or unwilling to hide their icons, and some provide an 
explicit notification that the app is in use, although this notification can 
usually be 
easily disabled in the Android notification settings.

\section{Related Work}
Previous studies \cite{stories_from_survivors, burke_using_2011}, the media 
\cite{shahani_smartphones_nodate, chatterjee_spyware_2018}, and IPV survivors 
\cite{ristenpart_ucsd_talk} have all reported 
that spyware is becoming an increasingly utilized means of performing
IPS. Tracking functions supported by spyware 
allow stalkers to monitor communications, track locations, and remotely 
activate cameras and microphones \cite{shahani_smartphones_nodate, 
chatterjee_spyware_2018}. Chatterjee et al. propose methods for searching for 
IPS-capable apps. Using Google search query expansion and unsupervised 
classification, they collected 9,224 apps possibly related to IPS on the Google 
Play store and the open Internet. They manually selected 70 apps for 
investigation. 
They observed the most fundamental functions of these apps are 
monitoring locations, communication logs and data, media contents, and phone 
usages. They also found that existing anti-spyware tools cannot effectively 
detect IPS-related apps \cite{chatterjee_spyware_2018}. In terms of previous 
work on app detection, mobile apps are often identified via the User-Agent 
string of the HTTP request, 
when one is available \cite{xu_identifying_2011}. With the increased use of 
HTTPS this may not always 
be the case, but we expect to find (based on conversations with authors of 
previous work) that security and best coding practices are not prioritized by 
stalkerware developers. Malware identification may be of use as well, 
since techniques are available to cluster malware that comes from the same 
source code into ``families'' \cite{hutchison_firma_2013}.
